---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## SBF: A R package for Shared Basis Factorization

Orthogonal Shared Basis Factorization (OSBF) is a joint matrix diagonalization
algorithm we developed for cross-species gene expression analysis.

### Installation

1. If you have git installed, clone `SBF` from Github and install.
```
git clone https://github.com/amalthomas111/SBF.git
```

If git is not installed, go to code -> Download ZIP. Unzip the file.

Then, inside an R console
```
install.packages("<path to SBF>/SBF", repos = NULL, type = "source")
```

[OR]

2. Install using `devtools`.
Install `devtools` if not already installed. Note: `devtools` requires lot of
dependencies.
```
# check devtools is installed. If not install.
pkgs <- c("devtools")
require_install <- pkgs[!(pkgs %in% row.names(installed.packages()))]
if (length(require_install))
  install.packages(require_install)
```

- Install SBF directly from Github via `devtools`
```
library(devtools)
devtools::install_github("amalthomas111/SBF")
```

Please contact us via e-mail or submit
a [GitHub issue](https://github.com/amalthomas111/SBF/issues)
if there is any trouble with installation.

### Quick demo

###  Load SBF package

```{r}
# load SBF package
library(SBF)
```

### Create a test dataset

- We will first create a test dataset. `SBF` package has the function 
  `createRandomMatrices` to create matrices with full column rank and
  different number of rows. All the matrices will have the same number of
  columns.
```{r}
set.seed(1231)
mymat <- createRandomMatrices(n = 4, ncols = 3, nrows = 4:6)
sapply(mymat, dim)
```
```{r}
mymat
```
Note: Depending upon the R versions, the random matrices generated could be
different.

The rank of the matrices
```{r}
sapply(mymat, function(x) {qr(x)$rank})
```

We will use the test dataset as our $D_i$ matrices.


### OSBF examples

The OSBF's shared orthogonal $V$, $U_i$'s with orthonormal columns, and
diagonal matrices $\Delta_i$'s can be estimated using the `SBF` function with
argument `orthogonal = TRUE`.

Check the arguments for the SBF function call using `?SBF`.

```{r}
# OSBF call
osbf <- SBF(matrix_list = mymat, orthogonal = TRUE)
```

```{r}
names(osbf)
```

For OSBF, the factorization error is optimized by default
(`minimizeError=TRUE`), and the `optimizeFactorization` function is invoked. 
Optimization could take some time depending on the data matrix (`mymat`) and
initial values.  
`?SBF` help function shows all arguments for the SBF call.

In OSBF, $V$ is orthogonal and columns of $U_i$'s are orthonormal
($U_i^T U_i = I$).

```{r}
zapsmall(t(osbf$v) %*% osbf$v)
```

```{r}
# check the columns of first matrix of U_ortho
zapsmall(t(osbf$u[[names(osbf$u)[[1]]]]) %*%
           osbf$u[[names(osbf$u)[[1]]]])
```

OSBF is not an exact factorization. The decomposition error is minimized and
the final factorization error is stored in `osbf$error`.
The initial factorization error we started with is stored in `osbf$error_start`.

```{r}
# initial decomposition error
osbf$error_start
# final decomposition error
osbf$error
```
The error decreased by a factor of `r round(osbf$error_start/osbf$error,2)`.

The number of iteration taken for optimizing:
```{r}
cat("For osbf, # iteration =", osbf$error_pos, "final error =", osbf$error)
```

A detailed explanation of the algorithm and example cases of factorization can be
found in the vignettes/docs directory.

### Factorization for an example gene expression dataset

- SBF package has a sample gene expression data with the mean expression of
nine tissues in five species. Now we will use this data as our $D_i$ matrices.

```{r}
# load sample dataset from SBF package
avg_counts <- SBF::TissueExprSpecies
# dimension of matrices for different species
sapply(avg_counts, dim)
# names of different tissue types in humans
names(avg_counts[["Homo_sapiens"]])
```

OSBF computation based on inter-sample correlation. 
$V$ is estimated using the mean $R_i^T R_i$.

```{r}
# OSBF call using correlation matrix
osbf_gem <- SBF(matrix_list = avg_counts, orthogonal = TRUE,
                transform_matrix = TRUE, tol = 1e-4)
# initial decomposition error
osbf_gem$error_start
# final decomposition error
osbf_gem$error
```
Note: For high-dimensional datasets, vary the tolerance threshold (`tol`) or
maximum number of iteration parameter (`max_iter`) to reduce the computing time.


When calling `SBF`, `orthogonal = FALSE` computes the SBF factorization,
an exact joint matrix factorization we developed. In SBF, the estimated
columns of $U_i$ are not orthonormal. More details and examples can be found
in the vignettes/docs directory.


### Contacts

***Amal Thomas*** *amalthom@usc.edu*


Contact us via e-mail or submit
a [GitHub issue](https://github.com/amalthomas111/SBF/issues)

### Publication

Available on [bioRxiv](https://biorxiv.org/cgi/content/short/2022.08.26.505467v1)

### Copyright

Copyright (C) 2018-2021  Amal Thomas

Authors: Amal Thomas

SBF is free software: you can redistribute it and/or modify it under
the terms of the GNU General Public License as published by the Free
Software Foundation, either version 3 of the License, or (at your
option) any later version.

SBF is distributed in the hope that it will be useful, but WITHOUT
ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
for more details.
