---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## SBF: A R package for Shared Basis Factorization

Shared Basis Factorization (SBF) is a joint matrix diagonalization
approach we developed for cross-species gene expression analysis.
Approximate Shared Basis Factorization (A-SBF) is an extension of the SBF
approach.

### Installation
- Clone from Github
```r
git clone https://github.com/amalthomas111/SBF.git
```
Inside an R console
```
library(devtools)
install("<path to SBF>/SBF")
```
- [OR] install directly from Github via `devtools`
```r
library(devtools)
devtools::install_github("amalthomas111/SBF")
# load package
library(SBF)
```

Please contact us via e-mail or through
a [GitHub issue](https://github.com/amalthomas111/SBF/issues)
if there is any trouble with installation.

### Quick demo

###  load SBF package

```{r}
# load SBF package
library(SBF)
```

### load test dataset

- SBF package has a sample gene expression data with the mean expression of
nine tissues in five species

```{r}
# load sample dataset from SBF package
avg_counts <- SBF::TissueExprSpecies
# dimension of matrices for different species
sapply(avg_counts, dim)
# names of different tissue types in humans
names(avg_counts[["Homo_sapiens"]])
```

### SBF computation

Estimating V using the sum of $D_i^T D_i$.
```{r}
# SBF call. Estimate V using the sum of Di^TDi
avg_counts <- SBF::TissueExprSpecies
sbf <- SBF(matrix_list = avg_counts)
```

`?SBF` help function shows all arguments for the SBF call.

```{r}
names(sbf)
```

Check whether the estimated $V$ is orthogonal.
```{r}
zapsmall(sbf$v %*% t(sbf$v))
```
Let us check the factorization error.

```{r}
# calculate decomposition error
decomperror <- calcDecompError(avg_counts, sbf$u, sbf$delta, sbf$v)
decomperror
```

Estimating V using sum of inverse-variance weighted $D_i^TD_i$.
```{r}
# SBF call. Estimate V using inverse-variance weighted Di^TDi
avg_counts <- SBF::TissueExprSpecies
sbf <- SBF(matrix_list = avg_counts, weighted = TRUE)
# calculate decomposition error
decomperror <- calcDecompError(avg_counts, sbf$u, sbf$delta, sbf$v)
decomperror
```

SBF computation based on inter-sample correlation. 
$V$ is estimated using $R_i^T R_i$.

```{r}
# SBF call using correlation matrix
avg_counts <- SBF::TissueExprSpecies
sbf_cor <- SBF(matrix_list = avg_counts, transform_matrix = TRUE)
decomperror <- calcDecompError(avg_counts, sbf_cor$u, sbf_cor$delta, sbf_cor$v)
decomperror
```

### Approximate-SBF (A-SBF)

Estimating V using the sum of $D_i^T D_i$ and estimating $U_i$'s
such that columns are orthonormal.
```{r}
# A-SBF call
avg_counts <- SBF::TissueExprSpecies
asbf <- SBF(matrix_list = avg_counts, approximate = TRUE)
```

```{r}
names(asbf)
```

In A-SBF, $V$ is orthogonal and columns of $U_i$'s are orthonormal
($U_i^T U_i = I$).

```{r}
zapsmall(t(asbf$v) %*% asbf$v)
```
```{r}
zapsmall(t(asbf$u_ortho[[names(asbf$u_ortho)[[1]]]]) %*%
           asbf$u_ortho[[names(asbf$u_ortho)[[1]]]])
```

A-SBF is not an exact factorization.

```{r}
# calculate decomposition error
decomperror <- calcDecompError(avg_counts, asbf$u_ortho, asbf$delta, asbf$v)
decomperror
```

This error is already computed and stored in `asbf$error`.


A-SBF call with inverse variance weighting. Estimating V using the sum of
inverse-variance weighted $D_i^T D_i$ and estimating $U_i$'s such that columns
are orthonormal.

```{r}
# A-SBF call with inverse variance weighting
avg_counts <- SBF::TissueExprSpecies
asbf_inv <- SBF(matrix_list = avg_counts, weighted = TRUE, approximate = TRUE)
# calculate decomposition error
decomperror_inv <- asbf_inv$error
decomperror_inv
```

A-SBF computation based on inter-sample correlation.
$V$ is estimated using $R_i^T R_i$.

````{r}
# A-SBF call using correlation matrix
avg_counts <- SBF::TissueExprSpecies
asbf_cor <- SBF(matrix_list = avg_counts, approximate = TRUE,
                transform_matrix = TRUE)
# calculate decomposition error
decomperror_cor <- asbf_cor$error
decomperror_cor
````
### Reduce A-SBF factorization error

Let us optimize the factorization for the three cases of A-SBF using the
`optimizeFactorization` function. Depending upon the initial values,
optimization could take some time.

```{r}
myopt <- optimizeFactorization(avg_counts, asbf$u_ortho, asbf$delta, asbf$v)
myopt_inv <- optimizeFactorization(avg_counts, asbf_inv$u_ortho, asbf_inv$delta,
                                   asbf_inv$v)
myopt_cor <- optimizeFactorization(avg_counts, asbf_cor$u_ortho, asbf_cor$delta,
                                   asbf_cor$v)
```

The number of iteration taken for optimizing and new factorization error:
```{r}
cat("For asbf, # iteration =", myopt$error_pos, "final error =", myopt$error)
cat("\nFor asbf inv, # iteration =", myopt_inv$error_pos, "final error =",
    myopt_inv$error)
cat("\nFor asbf cor, # iteration =", myopt_cor$error_pos, "final error =",
    myopt_cor$error)
```
```{r, echo = FALSE}
if ((round(myopt$error,2) == round(myopt_inv$error,2)) && (round(myopt$error,2) == round(myopt_cor$error,2))) {
  cat("For all three factorizations, after optimizing the final errors is the same(up to 2 decimals)")
  cat("\nThe final error is", round(myopt$error,2))  
}
```

### Contacts ###

***Amal Thomas*** *amalthom@usc.edu*

***Andrew D Smith*** *andrewds@usc.edu*

Contact us via e-mail or through
a [GitHub issue](https://github.com/amalthomas111/SBF/issues)

### Copyright ###

Copyright (C) 2018-2021  Amal Thomas and Andrew D. Smith

Authors: Amal Thomas and Andrew D. Smith

SBF is free software: you can redistribute it and/or modify it under
the terms of the GNU General Public License as published by the Free
Software Foundation, either version 3 of the License, or (at your
option) any later version.

SBF is distributed in the hope that it will be useful, but WITHOUT
ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
for more details.
