---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## SBF: A R package for Shared Basis Factorization

Approximate Shared Basis Factorization (A-SBF) is a joint matrix diagonalization
approach we developed for cross-species gene expression analysis.

### Installation
- Clone from Github
```r
git clone https://github.com/amalthomas111/SBF.git
```
Inside an R console
```
library(devtools)
install("<path to SBF>/SBF")
```
- [OR] install directly from Github via `devtools`
```r
library(devtools)
devtools::install_github("amalthomas111/SBF", build_vignettes = TRUE)
```

Please contact us via e-mail or submit
a [GitHub issue](https://github.com/amalthomas111/SBF/issues)
if there is any trouble with installation.

### Quick demo

###  Load SBF package

```{r}
# load SBF package
library(SBF)
```

### Analysis for a test dataset

- We will first create a test dataset. `SBF` package has the function 
  `createRandomMatrices` to create matrices with full column rank and
  different number of rows. All the matrices will have the same number of
  columns.
```{r}
set.seed(1231)
mymat <- createRandomMatrices(n = 4, ncols = 3, nrows = 4:6)
sapply(mymat, dim)
```

```{r}
sapply(mymat, function(x) {qr(x)$rank})
```

We will use the test dataset as our $D_i$ matrices.


### A-SBF examples

The A-SBF's shared orthogonal $V$, $U_i$'s with orthonormal columns and
diagonal matrices $\Delta_i$'s can be estimated using `SBF` function with
argument `approximate = TRUE`.

```{r}
# A-SBF call
asbf <- SBF(matrix_list = mymat, approximate = TRUE)
```


```{r}
names(asbf)
```

For A-SBF, the factorization error is optimized by default (`minimizeError=TRUE`)
and the `optimizeFactorization` function is invoked. 
Depending upon the data matrix (`mymat`) and initial values, optimization could
take some time. 
`?SBF` help function shows all arguments for the SBF call.

In A-SBF, $V$ is orthogonal and columns of $U_i$'s are orthonormal
($U_i^T U_i = I$).

```{r}
zapsmall(t(asbf$v) %*% asbf$v)
```

```{r}
# check the columns of first matrix of U_ortho
zapsmall(t(asbf$u[[names(asbf$u)[[1]]]]) %*%
           asbf$u[[names(asbf$u)[[1]]]])
```

A-SBF is not an exact factorization. The decomposition error is minimized and
stored in `asbf$error`.

```{r}
# initial decomposition error
asbf$error_start
# final decomposition error
asbf$error
```


The number of iteration taken for optimizing:
```{r}
cat("For asbf, # iteration =", asbf$error_pos, "final error =", asbf$error)
```

Detailed explanation of the math and examples case of factorization can be found
in the vignettes/docs directory.

### Factorization for an example gene expression dataset

- SBF package has a sample gene expression data with the mean expression of
nine tissues in five species. Now we will use this data as our $D_i$ matrices.

```{r}
# load sample dataset from SBF package
avg_counts <- SBF::TissueExprSpecies
# dimension of matrices for different species
sapply(avg_counts, dim)
# names of different tissue types in humans
names(avg_counts[["Homo_sapiens"]])
```

A-SBF computation based on inter-sample correlation. 
$V$ is estimated using the mean $R_i^T R_i$.

```{r}
# A-SBF call using correlation matrix
asbf_gem <- SBF(matrix_list = avg_counts, approximate = TRUE,
                transform_matrix = TRUE, tol = 1e-4)
# initial decomposition error
asbf_gem$error_start
# final decomposition error
asbf_gem$error
```
Note: For high-dimensional datasets, vary the tolerance threshold (`tol`) or
maximum number of iteration parameter (`max_iter`) to reduce the computing time.
<!-- For gene-expression analysis, if we want the shared space to represent -->
<!-- inter-sample correlation relationship, we do not update/change $V$ -->
<!-- while optimizing the factorization error. -->
<!-- In such cases, while reducing the factorization error -->
<!-- we set `optimizeV = FALSE` in the `optimizeFactorization` function. -->
<!-- We now optimize the factorization error to find the closest space. -->


When calling `SBF`, `approximate = FALSE` compute the SBF factorization which
is an exact join matrix factorization we developed. In SBF, the estimated
$U_i$'s are not orthonormal. More details and examples can be found
in the vignettes/docs directory.


### Contacts

***Amal Thomas*** *amalthom@usc.edu*

***Andrew D Smith*** *andrewds@usc.edu*

Contact us via e-mail or submit
a [GitHub issue](https://github.com/amalthomas111/SBF/issues)

### Copyright

Copyright (C) 2018-2021  Amal Thomas and Andrew D. Smith

Authors: Amal Thomas and Andrew D. Smith

SBF is free software: you can redistribute it and/or modify it under
the terms of the GNU General Public License as published by the Free
Software Foundation, either version 3 of the License, or (at your
option) any later version.

SBF is distributed in the hope that it will be useful, but WITHOUT
ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
for more details.
