---
title: "Gene expression analysis using A-SBF"
author: "Amal Thomas"
output:
  pdf_document:
    toc: yes
    number_section: yes
  knitr:::html_vignette:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
bibliography: references.bib
vignette: |
  %\VignetteIndexEntry{GeneExpressionAnalysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
text-align: justify}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{tr}}
\newcommand{\E}{\mathrm{E}}

# Introduction

Consider a set of $k$ real matrices $D_i \in \mathbb{R}^{m_i \times n}$,
each with full column rank.
Each $D_i$ represents the mean expression of $n$ tissues (columns) in species
$i$ with $m_i$ genes (rows) annotated for that species.
The $p\mbox{-th}$ column of each $D_i$ matrix corresponds to the average
gene-expression profile of a functionally equivalent tissue representing
a similar phenotype across species.
The number of rows is different for each $D_i$ matrix, but they all have the
same number of columns.

In A-SBF, we estimate an orthonormal basis $V$ for the common expression
space based on the inter-tissue correlations within each species.
Let $X_i \in \mathbb{R}^{m_i \times n}$ be the standardized gene
expression matrix for species $i$. We can write
\[
X_i = C_iD_i{S_i}^{-1},
\]
where
\[
C_i = I_{m_i} - {m_i}^{-1} 1_{m_i} {1_{m_i}}^T
\]
is a centering matrix and $S_i = \diag(s_1,\ldots,s_n)$ is a diagonal
scaling matrix, where $s_j$ is the standard deviation of $j$-th column
of $D_i$. Within species $i$, the correlation between expression
profiles for the $n$ tissues is
\[
R_i = X_i^TX_i/m_i.
\]
Each $R_i$ matrix has a dimension of $n \times n$ independent of the 
number of genes annotated in the species.
Since the corresponding column of each $D_i$ matrix represents a similar 
phenotype, we can define an expected correlation matrix across the
species:

\[
E = \E(R) = \textstyle\frac{1}{k}\sum_{i=1}^{k} R_i.
\]

The shared right basis matrix $V$ represents the common expression space
and is determined from the eigenvalue decomposition of $E$,
where $E=V \Lambda V^T$.
The right basis matrix $V$ is identical in all the $k$ matrix factorizations,
and the different dimensions of $V$ represent the correlation relationship
between tissues independent of the fact to which species tissues belong.
Once the $V$ is estimated, the species-specific left basis matrix with
orthonormal columns $U_i$ (eigengenes) and the $\Delta_i$ that minimize
the factorization error are computed 

```{r setup}
# load SBF package
library(SBF)
```

Additional package required for the vignette
```{r}
# install packages
pkgs <- c("data.table", "dplyr", "matrixStats")
require_install <- pkgs[!(pkgs %in% row.names(installed.packages()))]
if (length(require_install))
  install.packages(require_install)
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(matrixStats)
})
```


# Species and organs

In this workflow, we will work with gene expression profile of six tissues 
from eight species.
```{r}
species <- c("Homo_sapiens", "Pan_troglodytes", "Macaca_mulatta",
             "Mus_musculus", "Rattus_norvegicus", "Bos_taurus",
             "Sus_scrofa", "Gallus_gallus")
# function to create short name for species
getSpeciesShortName <- function(species_list) {
    require(data.table)
    shortname <- c()
    for (sp in species_list) {
       if (!grepl("_", sp)) {
          cat("\nspecies:", sp, "\n")
          stop("Invalid species full name")
       }
       shortname <- c(shortname,
                       paste0(tolower(substr(tstrsplit(sp, "_")[[1]], 1, 1)),
                       tstrsplit(sp, "_")[[2]]))
    }
    return(shortname)
}
species_short <- sapply(species, getSpeciesShortName)
species_short
# common tissues present in all 8 species
tissues <-  c("brain", "heart", "kidney", "liver", "lung", "testis")
```


# Load gene expression profiles

Download the processed RNA-Seq counts file
from https://figshare.com/s/4d8d5cf9c1362bcbe90d
```{r}
# set the path to the counts. Change it accordingly
counts_path <- "~/Dropbox/0.Analysis/0.paper/counts/"
counts_list <- metadata_list <- list()
require(data.table)
for (sp in species) {
  counts <- fread(paste0(counts_path, sp, "_logTPM.tsv"), sep = "\t",
                  header = T, data.table = FALSE, nThread = 4)
  row.names(counts) <- counts$V1
  counts$V1 <- NULL
  col_fields <- tstrsplit(colnames(counts), "_")
  metadata <- data.frame(
      project = col_fields[[1]],
      species = col_fields[[2]],
      tissue = col_fields[[3]],
      gsm = col_fields[[4]],
      name = colnames(counts),
      stringsAsFactors = F)
  metadata_sel <- metadata[metadata$tissue %in% tissues, , drop = F]
  counts_sel <- counts[, colnames(counts) %in% metadata_sel$name, drop = F]
  metadata_sel$ref <- seq_len(nrow(metadata_sel))
  metadata_sel$key <- paste0(metadata_sel$species, "_", metadata_sel$tissue)

  counts_list[[sp]] <- counts_sel
  metadata_list[[sp]] <- metadata_sel
}
```

The dimensions and the number of RNA-Seq profiles for different species.

```{r}
sapply(counts_list, dim)
```

Now, for each species, let us compute the average expression profile for each
tissue.

```{r}
#function to calculate mean expression values for each tissue/group
calcAvgCounts <- function(counts, metadata, ndecimal = 4) {
   require(dplyr)
   counts.avg <- list()
   if ("key" %in% colnames(metadata)) {
     for (species.organ in sort(unique(metadata$key))) {
        no.of.libs <- length(colnames(counts)[grepl(species.organ,
                                                    colnames(counts))])
        if (no.of.libs > 1)
           counts.avg[[species.organ]] <- round(rowMeans(counts[, colnames(counts)[
                                   grepl(species.organ, colnames(counts))]],
                                   dim = 1), ndecimal)
        else if (no.of.libs == 1)
           counts.avg[[species.organ]] <- round(counts[, colnames(counts)[grepl(
                              species.organ, colnames(counts))], drop = T], ndecimal)
     }
     avg.counts <- as.data.frame(counts.avg %>% bind_cols())
     row.names(avg.counts) <- row.names(counts)
     return(avg.counts)
   } else {
     stop("'key' column not found in the metadata!Exiting!")
   }
}

avg_counts <- list()
for (sp in species) {
    avg_counts[[sp]] <- calcAvgCounts(counts_list[[sp]],
                                      metadata_list[[sp]])
}
# check tissue columns are matching in each species
c_tissues <- as.data.frame(sapply(avg_counts, function(x) {
            tstrsplit(colnames(x), "_")[[2]]
  }))
if (!all(apply(c_tissues, 1, function(x) all(x == x[1])))) {
        stop("Error! tissues not matching")
}
```

The dimension of mean expression profiles
```{r}
sapply(avg_counts, dim)
```
Remove genes not expressed.

```{r}
removeZeros <- function(df) {
    return(df[rowSums(df) > 0, ])
}

avg_counts <- lapply(avg_counts, removeZeros)
sapply(avg_counts, dim)
# update counts_list
counts_list_sub <- list()
for (sp in names(avg_counts)) {
  counts_list_sub[[sp]] <- counts_list[[sp]][row.names(avg_counts[[sp]]), ,
                                                 drop = F]
}
```


# A-SBF

We will perform A-SBF in two ways.

1. Keeping the initial estimate of $V$ the same while updating $U_i$ and
$\Delta_i$ to minimize the factorization error.
By keeping the $V$ same, the initial $V$ estimated based inter-sample
correlation is maintained.
2. Update $V$, $U_i$ and $\Delta_i$ to minimize the factorization error.

```{r}
cat("\n=======================================================")
cat("\nASBF with optimizing V = FALSE started\n")
cat(format(Sys.time(), "%a %b %d %X %Y"), "\n")
t1 <- proc.time()
sbf_noVupdate <- SBF(avg_counts, transform_matrix = TRUE, approximate = TRUE,
                     optimizeV = FALSE, tol = 1e-3)#, tol = 1e-10)
t2 <- proc.time()
cat(format(Sys.time(), "%a %b %d %X %Y"), "\n")
cat("ASBF with optimizing V = FALSE finished\n")
cat("Time taken:\n")
t2 - t1
```


```{r}
cat("\n=======================================================")
cat("\nASBF with optimizing V=TRUE started\n")
cat(format(Sys.time(), "%a %b %d %X %Y"), "\n")
t1 <- proc.time()
sbf <- SBF(avg_counts, transform_matrix = TRUE, approximate = TRUE,
                     optimizeV = TRUE, tol = 1e-3)#, tol = 1e-10)
t2 <- proc.time()
cat(format(Sys.time(), "%a %b %d %X %Y"), "\n")
cat("ASBF with optimizing V=TRUE finished\n")
cat("Time taken:\n")
t2 - t1
```
Note: We use tolerance threshold = 1e-3 for fast computation. 
Reduce the tolerance threshold (default value 1e-10) to minimize 
the factorization error futher.
Lowering the threshold value increases the computing time.

The final factorization error and number of updates taken:
```{r}
cat("\n", sprintf("%-27s:", "Final error [No V update]"), sprintf("%16.2f",
                                                        sbf_noVupdate$error))
cat("\n", sprintf("%-27s:", "Final error [With V update]"), sprintf("%16.2f",
                                                                    sbf$error))
cat("\n", sprintf("%-27s:", "# of update [No V update]"), sprintf("%16d",
                                                      sbf_noVupdate$error_pos))
cat("\n", sprintf("%-27s:", "# of update [With V update]"), sprintf("%16d",
                                                                sbf$error_pos))
```
Optimization with updating $V$ achieves a lower decomposition error.

```{r}
sbf_noVupdate$error / sbf$error
```

Let us plot the decomposition error vs. updates.

```{r}
par(mfrow = c(1, 3))
plot(x = seq_len(length(sbf$error_vec)), y = sbf$error_vec,
     xlab = "# of updates (step 1 -> )",
     ylab = "Factorization error", col = "red")
plot(x = 10:length(sbf$error_vec),
     y = sbf$error_vec[10:length(sbf$error_vec)],
     xlab = "# of updates (step 10 -> )",
     ylab = "Factorization error", col = "red")
plot(x = 50:length(sbf$error_vec),
     y = sbf$error_vec[50:length(sbf$error_vec)],
     xlab = "# of updates (step 50 -> )",
     ylab = "Factorization error", col = "red")
```



Percentage of information ($p_{ij}$) represented by a common space dimension is
defined as
$\textstyle{p_{ij} = \delta_{ij}^2 / \sum_{j=1}^6 \delta_{ij}^2 \times 100}$,
$\mbox{ where } \textstyle{\Delta_i = \diag(\delta_{i1},\ldots, \delta_{i6})}$

```{r}
percentInfo_noVupdate <- lapply(sbf_noVupdate$d, function(x) {
  round(x^2 / sum(x^2) * 100, 2)})
cat("\nPercentage for each delta [No V update]:")
for (i in names(sbf_noVupdate$d)) {
  cat("\n", sprintf("%-25s:", i), sprintf("%8.2f", percentInfo_noVupdate[[i]]))
}
cat("\n-------------------------------------------------------------------------")


percentInfo <- lapply(sbf$d, function(x) {
  round(x^2 / sum(x^2) * 100, 2) })
cat("\nPercentage for each delta:")
for (i in names(sbf$d)) {
  cat("\n", sprintf("%-25s:", i), sprintf("%8.2f", percentInfo[[i]]))
}
```
The percentage of information represented by different dimensions of the two
approaches looks very similar.

# Project libraries into common space

Let us first compute $\Delta^{-1}$

```{r}
d_inv_NoVupdate <- list()
for (sp in names(avg_counts)) {
  if (length(sbf_noVupdate$d[[sp]]) == 1) {
    d_inv_NoVupdate[[sp]] <- as.matrix(diag(as.matrix(1 / sbf_noVupdate$d[[sp]])))
    } else {
      d_inv_NoVupdate[[sp]] <- as.matrix(diag(1 / sbf_noVupdate$d[[sp]]))
    }
}
d_inv <- list()
for (sp in names(avg_counts)) {
  if (length(sbf$d[[sp]]) == 1) {
    d_inv[[sp]] <- as.matrix(diag(as.matrix(1 / sbf$d[[sp]])))
    } else {
      d_inv[[sp]] <- as.matrix(diag(1 / sbf$d[[sp]]))
    }
}
```



Project individual profiles and average counts to common space by computing
$D_i^T U_i \Delta^{-1}$

```{r}
# project using no V update estimates
avgcounts_projected_noVupdate  <- counts_projected_noVupdate <- list()
n_noVupdate <- n1_noVupdate <- c()
for (sp in names(avg_counts)) {
  avgcounts_projected_noVupdate[[sp]] <- as.matrix(t(avg_counts[[sp]])) %*%
    as.matrix(sbf_noVupdate$u[[sp]]) %*% d_inv_NoVupdate[[sp]]
  n_noVupdate <- c(n_noVupdate, row.names(avgcounts_projected_noVupdate[[sp]]))
  counts_projected_noVupdate[[sp]] <- as.matrix(t(counts_list_sub[[sp]])) %*%
    as.matrix(sbf_noVupdate$u[[sp]]) %*% d_inv_NoVupdate[[sp]]
  n1_noVupdate <- c(n1_noVupdate, row.names(counts_projected_noVupdate[[sp]]))
}

# combine projected profiles
df_proj_avg_noVupdate <- as.data.frame(do.call(rbind, avgcounts_projected_noVupdate))
rownames(df_proj_avg_noVupdate) <- n_noVupdate
colnames(df_proj_avg_noVupdate) <- paste0("Dim", 1:ncol(df_proj_avg_noVupdate))
meta <- tstrsplit(row.names(df_proj_avg_noVupdate), "_")
df_proj_avg_noVupdate$tissue <- factor(meta[[2]])
df_proj_avg_noVupdate$species <- factor(meta[[1]])
df_proj_avg_noVupdate <- df_proj_avg_noVupdate %>% mutate(species = factor(species,
                            levels = species_short))


df_proj_noVupdate <- as.data.frame(do.call(rbind, counts_projected_noVupdate))
rownames(df_proj_noVupdate) <- n1_noVupdate
colnames(df_proj_noVupdate) <- paste0("Dim", 1:ncol(df_proj_noVupdate))
meta1 <- tstrsplit(row.names(df_proj_noVupdate), "_")
df_proj_noVupdate$tissue <- factor(meta1[[3]])
df_proj_noVupdate$species <- factor(meta1[[2]])
df_proj_noVupdate <- df_proj_noVupdate %>% mutate(species = factor(species,
                            levels = species_short))

# project using V update estimates
avgcounts_projected  <- counts_projected <- list()
n <- n1 <- c()
for (sp in names(avg_counts)) {
  avgcounts_projected[[sp]] <- as.matrix(t(avg_counts[[sp]])) %*%
    as.matrix(sbf$u[[sp]]) %*% d_inv[[sp]]
  n <- c(n, row.names(avgcounts_projected[[sp]]))
  counts_projected[[sp]] <- as.matrix(t(counts_list_sub[[sp]])) %*%
    as.matrix(sbf$u[[sp]]) %*% d_inv[[sp]]
  n1 <- c(n1, row.names(counts_projected[[sp]]))
}

# combine projected profiles
df_proj_avg <- as.data.frame(do.call(rbind, avgcounts_projected))
rownames(df_proj_avg) <- n
colnames(df_proj_avg) <- paste0("Dim", 1:ncol(df_proj_avg))
meta <- tstrsplit(row.names(df_proj_avg), "_")
df_proj_avg$tissue <- factor(meta[[2]])
df_proj_avg$species <- factor(meta[[1]])
df_proj_avg <- df_proj_avg %>% mutate(species = factor(species,
                            levels = species_short))


df_proj <- as.data.frame(do.call(rbind, counts_projected))
rownames(df_proj) <- n1
colnames(df_proj) <- paste0("Dim", 1:ncol(df_proj))
meta1 <- tstrsplit(row.names(df_proj), "_")
df_proj$tissue <- factor(meta1[[3]])
df_proj$species <- factor(meta1[[2]])
df_proj <- df_proj %>% mutate(species = factor(species,
                            levels = species_short))

```

# Clustering in the common space

```{r}
# install packages
pkgs <- c("ComplexHeatmap", "RColorBrewer")
require_install <- pkgs[!(pkgs %in% row.names(installed.packages()))]
if (length(require_install))
  install.packages(require_install)
suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(RColorBrewer)
})
```

Compute distances between projected profiles and perform clustering.

```{r}
data_noVupdate <- df_proj_noVupdate
data_noVupdate$tissue <- NULL
data_noVupdate$species <- NULL
data_noVupdate <- as.matrix(data_noVupdate)
data_noVupdate_dist <- as.matrix(dist(data_noVupdate, method = "euclidean"))
meta <- tstrsplit(colnames(data_noVupdate_dist), "_")
ht <- HeatmapAnnotation(tissue = meta[[3]], species = meta[[2]],
                        col = list(tissue = c("brain" = "#1B9E77",
                                              "heart" = "#D95F02",
                                              "kidney" = "#7570B3",
                                              "liver" = "#E7298A",
                                              "lung" = "#66A61E",
                                              "testis" = "#E6AB02"),
                                   species = c("hsapiens" = "#66C2A5",
                                               "ptroglodytes" = "#FC8D62",
                                               "mmulatta" = "#8DA0CB",
                                               "mmusculus" = "#E78AC3",
                                               "rnorvegicus" = "#A6D854",
                                               "btaurus" = "#FFD92F",
                                               "sscrofa" = "#E5C494",
                                               "ggallus" = "#B3B3B3")),
                        #show_annotation_name = F,
                        annotation_name_side = "left")
mypalette <- brewer.pal(9, "Blues")
morecolors <- colorRampPalette(mypalette)

myheatmap1 <- Heatmap(as.matrix(data_noVupdate_dist), cluster_rows = T,
                     clustering_method_rows = "centroid",
                     cluster_columns = T,
                     clustering_method_columns = "centroid",
                     top_annotation = ht, col = morecolors(50),
                     show_row_names = F, show_column_names = F,
                     name = "distance")
myheatmap1
```

Compute distances between projected profiles in the common space estimated
using optimized $V$ and perform clustering.
```{r}
data <- df_proj
data$tissue <- NULL
data$species <- NULL
data <- as.matrix(data)
data_dist <- as.matrix(dist(data, method = "euclidean"))
meta <- tstrsplit(colnames(data_dist), "_")
ht <- HeatmapAnnotation(tissue = meta[[3]], species = meta[[2]],
                        col = list(tissue = c("brain" = "#1B9E77",
                                              "heart" = "#D95F02",
                                              "kidney" = "#7570B3",
                                              "liver" = "#E7298A",
                                              "lung" = "#66A61E",
                                              "testis" = "#E6AB02"),
                                   species = c("hsapiens" = "#66C2A5",
                                               "ptroglodytes" = "#FC8D62",
                                               "mmulatta" = "#8DA0CB",
                                               "mmusculus" = "#E78AC3",
                                               "rnorvegicus" = "#A6D854",
                                               "btaurus" = "#FFD92F",
                                               "sscrofa" = "#E5C494",
                                               "ggallus" = "#B3B3B3")),
                        #show_annotation_name = F,
                        annotation_name_side = "left")
mypalette <- brewer.pal(9, "Blues")
morecolors <- colorRampPalette(mypalette)

myheatmap <- Heatmap(as.matrix(data_dist), cluster_rows = T,
                     clustering_method_rows = "centroid",
                     cluster_columns = T,
                     clustering_method_columns = "centroid",
                     top_annotation = ht, col = morecolors(50),
                     show_row_names = F, show_column_names = F,
                     name = "distance")
myheatmap
```

Gene expression profiles cluster by tissue type independent of the species of
origin.


# Two-dimensional projection plots in the common space

Next, we will explore the 2D projection plots in the common space. We will first
define a custom theme that we will use for the plots.

```{r}
# install packages
pkgs <- c("grid", "ggthemes", "ggplot2")
require_install <- pkgs[!(pkgs %in% row.names(installed.packages()))]
if (length(require_install))
  install.packages(require_install)
suppressPackageStartupMessages({
  library(grid)
  library(ggthemes)
  library(ggplot2)
})
```

We will use the following custom theme for ggplots.
```{r}
# custom theme function for ggplot2
customTheme <- function(base_size = 10, base_family="helvetica") {
   require(grid)
   require(ggthemes)
   (theme_foundation(base_size = base_size)
   + theme(plot.title = element_text(face = "bold",
                                   size = rel(1.2), hjust = 0.5),
         text = element_text(),
         panel.background = element_rect(colour = NA),
         plot.background = element_rect(colour = NA),
         panel.border = element_rect(colour = NA),
         axis.title = element_text(size = rel(1)),
         axis.title.y = element_text(angle = 90, vjust = 2),
         axis.title.x = element_text(vjust = -0.2),
         axis.text = element_text(),
         axis.line = element_line(colour = "black"),
         axis.ticks = element_line(),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         legend.key = element_rect(colour = NA),
         legend.position = "top",
         legend.direction = "horizontal",
         legend.key.size = unit(0.2, "cm"),
         legend.spacing = unit(0, "cm"),
         legend.title = element_text(face = "italic"),
         plot.margin = unit(c(10, 5, 5, 5), "mm"),
         strip.background = element_rect(colour = "#f0f0f0", fill = "#f0f0f0"),
         strip.text = element_text(face = "bold")))
}
```


Let us first check the projected libraries in dimension 1 and 2.
```{r}
# 2D plot for Dim1 and Dim2 [No V update]
sel_colors <- brewer.pal(8, "Dark2")[seq_len(length(unique(df_proj_noVupdate$tissue)))]
i <- 1
j <- 2
ggplot(df_proj_noVupdate, aes(x = df_proj_noVupdate[, i],
                              y = df_proj_noVupdate[, j], col = tissue,
                              shape = species, fill = tissue)) +
  xlab(paste("A-SBF Dim", i)) + ylab(paste("A-SBF Dim", j)) +
  geom_point(size = 1.5) + scale_color_manual(values = sel_colors) +
  scale_shape_manual(values = c(21:25, 3:7)) +
  scale_fill_manual(values = sel_colors) +
  customTheme() +
  theme(legend.title = element_blank())
```

The same with optimized V
```{r}
# 2D plot for Dim1 and Dim2 [With V update]
sel_colors <- brewer.pal(8, "Dark2")[seq_len(length(unique(df_proj$tissue)))]
i <- 1
j <- 2
ggplot(df_proj, aes(x = df_proj[, i],
                              y = df_proj[, j], col = tissue,
                              shape = species, fill = tissue)) +
  xlab(paste("A-SBF Dim", i)) + ylab(paste("A-SBF Dim", j)) +
  geom_point(size = 1.5) + scale_color_manual(values = sel_colors) + 
  scale_shape_manual(values = c(21:25, 3:7)) +
  scale_fill_manual(values = sel_colors) +
  customTheme() +
  theme(legend.title = element_blank())
```

Let us plot the projected libraries in dimensions 2 and 3.

```{r}
# 2D plot for Dim1 and Dim2 [No V update]
sel_colors <- brewer.pal(8, "Dark2")[seq_len(length(unique(df_proj_noVupdate$tissue)))]
i <- 2
j <- 3
ggplot(df_proj_noVupdate, aes(x = df_proj_noVupdate[, i],
                              y = df_proj_noVupdate[, j], col = tissue,
                              shape = species, fill = tissue)) +
  xlab(paste("A-SBF Dim", i)) + ylab(paste("A-SBF Dim", j)) +
  geom_point(size = 1.5) + scale_color_manual(values = sel_colors) + 
  scale_shape_manual(values = c(21:25, 3:7)) +
  scale_fill_manual(values = sel_colors) +
  customTheme() +
  theme(legend.title = element_blank())
```

```{r}
# 2D plot for Dim1 and Dim2 [With V update]
sel_colors <- brewer.pal(8, "Dark2")[seq_len(length(unique(df_proj$tissue)))]
i <- 2
j <- 3
ggplot(df_proj, aes(x = df_proj[, i], y = df_proj[, j], col = tissue,
                              shape = species, fill = tissue)) +  
  xlab(paste("A-SBF Dim", i)) + ylab(paste("A-SBF Dim", j)) +
  geom_point(size = 1.5) + scale_color_manual(values = sel_colors) + 
  scale_shape_manual(values = c(21:25, 3:7)) +
  scale_fill_manual(values = sel_colors) +
  customTheme() +
  theme(legend.title = element_blank())
```

Similarly, we can check for other dimensions of the common space.
We observe that both clustering and 2D projections plots in the common space
with optimized V are similar to those without V update.
So we will use the common space with optimized V for future analysis.


# Expression specificity and eigengene loadings

Functions to compute tissue specificity ($\tau$) and scaled average
expression profile.

```{r}
# function to compute Tau
calc_tissue_specificity <- function(a) {
    a <- as.matrix(a)
    b <- a / rowMaxs(a)
    return(rowSums(1 - b) / (ncol(b) - 1))
}
    
Tau <- lapply(avg_counts, function(x) { calc_tissue_specificity(x)})
avg_counts_scaled <- lapply(avg_counts, function(x) { t(scale(t(x)))})

combine_expr <- list()
for (sp in names(avg_counts_scaled)) {
  x <- as.data.frame(avg_counts_scaled[[sp]])
  x[["Tau"]] <- Tau[[sp]]
  combine_expr[[sp]] <- x
}
```

From the 2D projection plot, we find that along the dimension 2, testis profiles
lie on the negative axis and separate from the rest of the tissues.
We will plot the relationship between $U_i$ loadings in dimension 2 and
the testis expression.


```{r}
sel_dim <- 2
sel_tissue <- "testis"
species <- "Homo_sapiens"
expr <- combine_expr[[species]]
asbf_coef <- sbf$u[[species]]
expr[["coef"]] <- asbf_coef[, sel_dim, drop = T]
expr1 <- expr[, c(paste0(getSpeciesShortName(species), "_", sel_tissue),
                   "Tau", "coef")]
colnames(expr1) <- c("tissue_zscore", "Tau", "coef")
head(expr1)
```

Dimension 2 $U_i$ loadings vs expression specificity ($\tau$) for humans

```{r}
# function to return scientific name
species_scientific <- function(x) {
  parts <- tstrsplit(x, "_")
  return(paste0(substr(x, 1, 1), ".", parts[[2]]))
}

# plot scatter
mid <- 0
p1 <- ggplot(expr1, aes(x = Tau, y = coef, col = tissue_zscore)) + theme_bw() +
  geom_point(size = 0.5) + xlab("Expression specificity") +
  ylab(paste0("Dim", sel_dim, " Coefficient")) +
  scale_color_gradient2(midpoint = mid, low = "blue", mid = "white",
                        high = "red", space = "Lab") +
  scale_y_continuous(limits = c(-1 * max(abs(expr1$coef)), max(abs(expr1$coef))),
                     breaks = seq(-1 * round(max(abs(expr$coef)), 2),
                                  round(max(abs(expr$coef)), 2), by = 0.01)) +
  customTheme() +  theme(legend.position = "right",
                         legend.direction = "vertical") +
  labs(title = paste0(species_scientific(species)), color = "Z-score") +
  theme(legend.key.size = unit(0.5, "cm"),
        plot.title = element_text(face = "italic"))
p1
```

Dimension 2 $U_i$ loadings vs expression specificity ($\tau$) for pig.


```{r}
sel_dim <- 2
sel_tissue <- "testis"
species <- "Sus_scrofa"
expr <- combine_expr[[species]]
asbf_coef <- sbf$u[[species]]
expr[["coef"]] <- asbf_coef[, sel_dim, drop = T]
expr1 <- expr[, c(paste0(getSpeciesShortName(species), "_", sel_tissue),
                   "Tau", "coef")]
colnames(expr1) <- c("tissue_zscore", "Tau", "coef")

# plot scatter
mid <- 0
p2 <- ggplot(expr1, aes(x = Tau, y = coef, col = tissue_zscore)) + theme_bw() +
  geom_point(size = 0.5) + xlab("Expression specificity") +
  ylab(paste0("Dim", sel_dim, " Coefficient")) +
  scale_color_gradient2(midpoint = mid, low = "blue", mid = "white",
                        high = "red", space = "Lab") +
  scale_y_continuous(limits = c(-1 * max(abs(expr1$coef)), max(abs(expr1$coef))),
                     breaks = seq(-1 * round(max(abs(expr$coef)), 2),
                                  round(max(abs(expr$coef)), 2), by = 0.01)) +
  customTheme() +  theme(legend.position = "right",
                         legend.direction = "vertical") +
  labs(title = paste0(species_scientific(species)), color = "Z-score") +
  theme(legend.key.size = unit(0.5, "cm"),
        plot.title = element_text(face = "italic"))
p2
```

# GO analysis

Download the GO files from: https://figshare.com/s/d96c586d5e53199d5370
<!-- 10.6084/m9.figshare.19633131 -->
We will perform the gene ontology analysis for genes with high coefficients.
For GO analysis, we will use shared 1-to-1 orthologs in eight species as
the background.

```{r}
# load GO analysis files
# go_path <- "path to shared 1-to-1 orthologs"
go_path <- "~/Dropbox/0.Analysis/0.paper/GOKeggFiles/"
file <- "allwayOrthologs_hsapiens-ptroglodytes-mmulatta-mmusculus-rnorvegicus-btaurus-sscrofa-ggallus_ens94.tsv"
orthologs <- read.table(file.path(go_path, file), header = T, sep = "\t")
head(orthologs)
```

We will use the goseq Bioconductor package to perform GO enrichment analysis.


```{r}
# install packages
pkgs <- c("goseq")
require_install <- pkgs[!(pkgs %in% row.names(installed.packages()))]
if (length(require_install)) {
  if (!require("BiocManager", quietly = TRUE))
      install.packages("BiocManager")
  BiocManager::install("goseq")
}
suppressPackageStartupMessages({
  library(goseq)
})
```

Let us perform GO analysis for top testis-specific genes in Dimension 2.
The testis-specific genes have negative loadings.


```{r}
sel_dim <- 2
sel_tissue <- "testis"
top_genes <- 100
sel_sign <- "neg"
```

We will perform GO analysis for human first.

```{r}
species <- "Homo_sapiens"
expr <- combine_expr[[species]]
asbf_coef <- sbf$u[[species]]
expr[["coef"]] <- asbf_coef[, sel_dim, drop = T]
expr1 <- expr[, c(paste0(getSpeciesShortName(species), "_", sel_tissue), "Tau",
                  "coef")]
colnames(expr1) <- c("tissue_zscore", "Tau", "coef")
if (sel_sign == "neg") {
  cat("\n selecting negative loadings")
  expr1_selsign <- expr1[expr1$coef < 0, ]
} else {
  cat("\n selecting positive loadings")
  expr1_selsign <- expr1[expr1$coef >= 0, ]
}
expr1_selsign$score <- expr1_selsign$Tau * abs(expr1_selsign$coef)
expr1_selsign$rank <- rank(-1 * expr1_selsign$score)
expr1_selsign <- expr1_selsign[order(expr1_selsign$rank), ]
genes_fg <- row.names(expr1_selsign[expr1_selsign$rank <= top_genes, ])
if (species == "Homo_sapiens") {
  cat("\nUsing orthologs (human IDs) as background")
  genes_bg <- orthologs$hsapiens
}
if (species == "Mus_musculus") {
  cat("\nUsing orthologs (mouse IDs) as background")
  genes_bg <- orthologs$mmusculus
}
genes_bg <- genes_bg[!genes_bg %in% genes_fg]
genome <- "hg38"
total_genes <- unique(c(genes_fg, genes_bg))
up_genes <- as.integer(total_genes %in%  genes_fg)
names(up_genes) <- total_genes
```


```{r, echo = FALSE, warning = FALSE}
# load("hg38 length data")
load("~/Dropbox/0.Analysis/0.paper/GOKeggFiles/hg38_length.EnsemblV94.RData")
lengthData.up <- lengthData[names(up_genes)]
# load("hg38 EnsembleID to GO data")
load("~/Dropbox/0.Analysis/0.paper/GOKeggFiles/hg38_geneID2GO.EnsemblID2GO.EnsembleV94.Robj")
pwf <- nullp(up_genes, bias.data = lengthData.up, plot.fit = F)
go <- goseq(pwf, "hg38", "ensGene", gene2cat = geneID2GO, test.cats = c("GO:BP"))
go.sub <- go[go$ontology == "BP", ]
go.sub$padj <- p.adjust(go.sub$over_represented_pvalue, method = "BH")
go.sub[["ratio"]] <- round(go.sub[["numDEInCat"]] / go.sub[["numInCat"]], 4)
go.sub <- go.sub[with(go.sub, order(padj, decreasing = c(FALSE))), ]
go.sub$over_represented_pvalue <- NULL
go.sub$under_represented_pvalue <- NULL
```

```{r}
head(go.sub)
```

Barplot with top human GO terms and their p-value.

```{r}
# GO enrichment plot for human
go_out <- head(go.sub, n = 8)
go_out$padj <- as.numeric(go_out$padj)
go_out$term <- factor(go_out$term, levels = go_out$term)
breaks <- round(c(0, 1 / 4, 2 / 4, 3 / 4, 1) * max(go_out[["ratio"]]), 2)
go_plot <- ggplot(go_out, aes(x = term, y = ratio, fill = padj)) + geom_col() +
  scale_y_continuous(expand = c(0, 0), breaks = breaks,
                     limits = c(0, max(go_out[["ratio"]] + 0.05))) +
  scale_x_discrete() + coord_flip() +
  scale_color_gradient(low = "blue", high = "red") +
  ylab(paste0("Ratio of genes in GO category (",
              species_scientific(species), ")")) +
  xlab("") + customTheme() + theme(legend.position = "right",
                                   legend.direction = "vertical",
                                   plot.margin = unit(c(10, 5, 5, 5), "mm"))
go_plot
```


GO analysis for mouse.

```{r}
species <- "Mus_musculus"
expr <- combine_expr[[species]]
asbf_coef <- sbf$u[[species]]
expr[["coef"]] <- asbf_coef[, sel_dim, drop = T]
expr1 <- expr[, c(paste0(getSpeciesShortName(species), "_", sel_tissue), "Tau",
                  "coef")]
colnames(expr1) <- c("tissue_zscore", "Tau", "coef")
if (sel_sign == "neg") {
  cat("\n selecting negative loadings")
  expr1_selsign <- expr1[expr1$coef < 0, ]
} else {
  cat("\n selecting positive loadings")
  expr1_selsign <- expr1[expr1$coef >= 0, ]
}
expr1_selsign$score <- expr1_selsign$Tau * abs(expr1_selsign$coef)
expr1_selsign$rank <- rank(-1 * expr1_selsign$score)
expr1_selsign <- expr1_selsign[order(expr1_selsign$rank), ]
genes_fg <- row.names(expr1_selsign[expr1_selsign$rank <= top_genes, ])
if (species == "Homo_sapiens") {
  cat("\nUsing orthologs (human IDs) as background")
  genes_bg <- orthologs$hsapiens
}
if (species == "Mus_musculus") {
  cat("\nUsing orthologs (mouse IDs) as background")
  genes_bg <- orthologs$mmusculus
}
genes_bg <- genes_bg[!genes_bg %in% genes_fg]
genome <- "mm10"
total_genes <- unique(c(genes_fg, genes_bg))
up_genes <- as.integer(total_genes %in%  genes_fg)
names(up_genes) <- total_genes
```

```{r, echo = FALSE, warning = FALSE}
# load("mm10 length data")
load("~/Dropbox/0.Analysis/0.paper/GOKeggFiles/mm10_length_EnsemblV94.RData")
lengthData.up <- lengthData[names(up_genes)]
# load("mm10 EnsembleID to GO data")
load("~/Dropbox/0.Analysis/0.paper/GOKeggFiles/mm10_geneID2GO.EnsemblID2GO.EnsembleV94.Robj")
pwf <- nullp(up_genes, bias.data = lengthData.up, plot.fit = F)
go <- goseq(pwf, "mm10", "ensGene", gene2cat = mouse_geneID2GO, test.cats = c("GO:BP"))
go.sub <- go[go$ontology == "BP", ]
go.sub$padj <- p.adjust(go.sub$over_represented_pvalue, method = "BH")
go.sub[["ratio"]] <- round(go.sub[["numDEInCat"]] / go.sub[["numInCat"]], 4)
go.sub <- go.sub[with(go.sub, order(padj, decreasing = c(FALSE))), ]
go.sub$over_represented_pvalue <- NULL
go.sub$under_represented_pvalue <- NULL
```

```{r}
head(go.sub)
```

Barplot with top mouse GO terms and their p-value.

```{r}
# GO enrichment plot for mouse
go_out <- head(go.sub, n = 8)
go_out$padj <- as.numeric(go_out$padj)
go_out$term <- factor(go_out$term, levels = go_out$term)
breaks <- round(c(0, 1 / 4, 2 / 4, 3 / 4, 1) * max(go_out[["ratio"]]), 2)
go_plot <- ggplot(go_out, aes(x = term, y = ratio, fill = padj)) + geom_col() +
  scale_y_continuous(expand = c(0, 0), breaks = breaks,
                     limits = c(0, max(go_out[["ratio"]] + 0.05))) +
  scale_x_discrete() + coord_flip() +
  scale_color_gradient(low = "blue", high = "red") +
  ylab(paste0("Ratio of genes in GO category (",
              species_scientific(species), ")")) +
  xlab("") + customTheme() + theme(legend.position = "right",
                                   legend.direction = "vertical",
                                   plot.margin = unit(c(10, 5, 5, 5), "mm"))
go_plot
```

